---
title: "Data Analysis Report - your title"
author: "Elizabeth Yirava, Joanna Giju, Nicholas Rohrs"
date: "04/17/2024"
format:
  pdf:
    documentclass: article
    geometry:
      - margin=1in
    fontsize: 12pt
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r}
#| label: setup

## Good practice to include libraries and functions you will need in a chunk
## at the beginning of your file
library(jtools)
library(patchwork)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(broom)
library(here)
## functions
logit <- function(x) log(x/(1 - x))
invlogit <- function(x) exp(x)/(1 + exp(x))

## This sets the gpplot theme to white for all plots
## the option `base_size` sets the base font size for all plots
theme_set(
    theme_bw(base_size = 8)
)
```

# Abstract

Write a summary of your project (around 300 words) here. This section should provide a concise overview of your research question, data, methods, and key findings.

# Introduction

In this section, introduce the problem and question(s) of interest that you aim to address in your analysis. Provide relevant background information and explain the motivation and significance of your research question. Cite manuscripts or websites you use. Here an example of how to cite a paper, @wickham2019. See  @quarto_cit for how to manage citations in Quarto.


# Data and Methods

Describe the data you used in your analysis, including the source, characteristics, and any pre-processing or cleaning steps you performed. 

```{r}
# Option 1: tidytuesdayR package 
## install.packages("tidytuesdayR")

tuesdata <- tidytuesdayR::tt_load('2023-12-05')

life_expectancy <- tuesdata$life_expectancy
life_expectancy_different_ages <- tuesdata$life_expectancy_different_ages
life_expectancy_female_male <- tuesdata$life_expectancy_female_male

```



```{r}
working_dir <- here::here()


le <- read_csv("life_expectancy.csv")
colnames(le)[4] <- "LifeExpectancy"
readr::write_csv(
  le,
  fs::path(working_dir, "life_expectancy.csv")
)

le_different_ages <- read_csv("life_expectancy_different_ages.csv")
colnames(le_different_ages)[4:9] <- c("LifeExpectancy0", "LifeExpectancy10", "LifeExpectancy25","LifeExpectancy45", "LifeExpectancy65", "LifeExpectancy80")
readr::write_csv(
  le_different_ages,
  fs::path(working_dir, "life_expectancy_different_ages.csv")
)

le_female_male <- read_csv("life_expectancy_female_male.csv")
colnames(le_female_male)[4] <- "LifeExpectancyDiffFM"
readr::write_csv(
  le_female_male,
  fs::path(working_dir, "life_expectancy_female_male.csv")
```

```{r}
# Add a new column named 'Past62' indicating whether individuals are expected to live past 62
le$Past62 <- ifelse(le$LifeExpectancy > 62, 1, 0)

# List of countries to include
countries_to_include <- c("United States", "China", "Japan", "Germany", "India", "United Kingdom", "France")

# Filter the dataset to include only the specified countries
le_filtered <- le %>% filter(Entity %in% countries_to_include)
```

This is a good place where to include results of your exploratory data analysis. See @fig-eda as an example. 
Check the .qmd file for an example of how to include plots, control their size and include references in the text. 


You should also explain the methods you employed. I expect you to define the statistical model you chose for your data in mathematical form, as we did throughout the course. 

Crab example: we define as outcome the presence/absence of a satellite for each crab and denote the observed values as $y_i$ with $i = 1, \ldots, n$ with $n = 173$. Let $y_i = 1$ indicate presence of a satellite and $0$ otherwise. 

We assume that each $y_i$ is a realization of a random variables $Y_i \sim \mbox{Bernoulli}(p_i)$ independently, and model the probability of having a satellite for each crab using the following logistic regression:

$$
\begin{aligned}
  \mbox{logit}(p_i) = & \beta_0 + \\ 
                    & \beta_1 \times \text{I(color == medium)}_i  \\
                   & \beta_2 \times \text{I(color == dark medium)}_i \\
                   & \beta_3 \times \text{I(color == dark)}_i \\
                   & \beta_4  \times \text{I(width)}_i
\end{aligned}
$$ {#eq-crab-model}

where $\mbox{I}()$ in @eq-crab-model indicates a dummy variable, taking value 1 when the condition is true and zero otherwise. 

# Results
Present your results in this section, including figures, tables, and any other relevant outputs from your analysis. 

```{r}
# GLM for the proportion of individuals expected to live beyond 62

binomial_model_country <- glm(Past62 ~ Entity, data = le_filtered, family = binomial)
summary(binomial_model_country)
```

```{r}
# Proportion of those expected to live beyond 62 in each country
proportion_past_62_by_country <- le %>%
  group_by(Entity) %>%
  summarize(proportion_past_62 = mean(Past62))
```

```{r}
# Proportion expected to live beyond 62 in XYZ vs proportion expected to live beyond 62 in the US divided by decades
proportion_past_62_us <- proportion_past_62_by_country %>%
  filter(Entity == "United States")

le_decades <- le %>%
  mutate(Decade = as.integer(Year %/% 10) * 10) %>%
  group_by(Entity, Decade) %>%
  summarize(proportion_past_62 = mean(Past62))
```

```{r}
# Life expectancy changing across US decades
le_us_decades <- le %>%
  filter(Entity == "United States") %>%
  mutate(Decade = as.integer(Year %/% 10) * 10) %>%
  group_by(Decade) %>%
  summarize(proportion_past_62 = mean(Past62))
```

```{r}
# Gender difference in life expectancy over years
```

```{r}
# Line chart showing female/male difference across the years included in data
```

```{r}
# Line chart showing 6 lines (one for each year of life expectancy 0-80) across all years included in data
le_lines <- le %>%
  pivot_longer(cols = starts_with("LifeExpectancy"), names_to = "Age", values_to = "LifeExpectancy") %>%
  mutate(Decade = as.integer(Year %/% 10) * 10) %>%
  ggplot(aes(x = Year, y = LifeExpectancy, color = Age)) +
  geom_line() +
  labs(title = "Life Expectancy Over Time by Age", x = "Year", y = "Life Expectancy", color = "Age")
plot(le_lines)
```

```{r}
# Mosaic plot with proportions of those living past age 62 vs those who don’t for each year
```

To display the model summary in a nice format you can use the `summ()` function from the `jtools` package.
 
Other libraries: 
\begin{itemize}
\item \texttt{xtable}: good if you are familiar with latex (see \href{https://cran.r-project.org/web/packages/xtable/vignettes/xtableGallery.pdf}{xtable documentation} for examples)
\item \texttt{gtable}, implements a ``grammar of table'' idea (see \url{https://gt.rstudio.com} for documentation)
\end{itemize}

In this section, discuss and interpret your results. Explain the implications and significance of your findings, and relate them to your research question and existing literature or theories.

# Conclusion

Summarize your main findings and conclusions in this section. Highlight the key takeaways and contributions of your analysis, and provide recommendations or suggestions for future research or practical applications based on your results.

### References

::: {#refs}
:::

# Appendix (Optional)

If you want to include your R code as an appendix, you can create a new code chunk and set `#| echo: true` to show the code and `#| eval: false` to avoid that to be run. You can also provide scripts separately as supplementary material.

```{r}
#| echo: true
#| eval: false

# Your code here
```


